TO FIND

### 2013-05-30 ms.tex

\textbf{Title:}  Using neutral theory to reveal the contribution of dispersal to community assembly in complex landscapes\\

@@

%========================================================%
\section*{Abstract}
The metacommunity perspective appears naturally as the appropriate conceptual
framework to make ecology more predictive, integrating elements of theoretical
ecology. The recent development of neutral theory appears as a step back in that
direction because of the assumption of ecological equivalence and the absence of
any effect of the environment on community organization. A remarkable strength
of neutral theory is nonetheless to provide a general theory of diversity that
accounts for a wide range of empirical observations. In this paper, we argue
that neutral theory can be a useful tool to understand the impact of dispersal
on community organization in landscapes of various complexities. Our main
objective is to use neutral theory to stress the importance of complex landscape
network structure on the distribution of diversity. We refer to the landscape
organization as a "spatial contingency" that could potentially affect the
coexistence mechanisms at play. We briefly review the main approaches to
describe spatial networks and describe three simple toy models of metacommunity
dynamics. We take this opportunity to review their assumptions and main
predictions. We then conduct simple simulations of these models to reveal with
simple examples the impact of spatial network structure on diversity
distribution. The simulation results show that competitive interactions buffer
the potential impact of landscape structure. The strongest centrality-species
richness relationship was observed for the patch dynamics, a model without any
interactions. On the other hand, strong and unequal competitive interactions
minimized the effect of centrality. We conclude that the neutral model is thus a
useful tool to understand the joint effects of dispersal and community
interactions. Our analysis shows that ecologists must now integrate more
realistic landscapes when analyzing community assembly from a metacommunity
perspective.
\newpage

@@

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2012}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of details they put in, leaving out some elements they consider
unimportant. The realism perspective requires that all assumptions of the theory
must be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Shmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2012} is a satisfying approximation.

@@

To a vast extent, these relationships are preserved when looking at the
geodesic distances (Fiig. 4), i.e. along how many edges should one travel to
connect two patchs. Interestingly enough, the distance-dissimilarity
relationship for the neutral model is markedly hump-shaped, with sites being at
a medium distance having the maximal dissimilarity.

@@

The network approach to the study of spatially explicit landscapes is a major
advancement in metacommunity ecology. It is a first step to make the concept
operational because it accounts for more realistic landscape structures and
dispersal kernels. It makes a significant departure to simple island-mainland or
global dispersal approaches used previously (e.g. \parencite{Tilman1994a,
Hubbell2001, Mouquet2002}). But dispersal is also spatially explicit in a
lattice model and it does not make the landscape more realistic. We believe the
fundamental contribution of this approach is the consideration of spatial
heterogeneity of dispersal. In agreement with previous theoretical
\parencite{Economo2011, Desjardins2012b} and experimemental studies
\parencite{Carrara2012}, the simulations show that the degree centrality has a
significant impact on $\alpha$ diversity. Central nodes might also contribute
more to maintain $\gamma$ diversity, as they are essential for species to spread
throughout the landscape. The nodes could be potentially quantified as keystone
for the metacommunity \parencite{Mouquet2013}. Interestingly, but not
surprisingly, this effect is weaker with species sorting dynamics. We could even
hypothesize it will vanish in the extreme case of niche differentiation (with
low overlap for instance) and low mass effect. In this particular case, the
neutral versus niche comparison therefore illustrates that very strong unequal
competitive interactions could overwhelm the impact of dispersal.

>>>>

\textbf{Title:} Using neutral theory to reveal the contribution of meta-community processes to assembly in complex landscapes\\

@@

%========================================================%
\section*{Abstract}
The metacommunity perspective appears as an appropriate conceptual
framework to make ecology more predictive. The recent development of neutral theory appears as a step back in that
direction because of the assumption of ecological equivalence and the absence of
any effect of the environment on community organization. A remarkable strength
of neutral theory is nonetheless to provide a general theory of diversity that
accounts for a wide range of empirical observations. In this paper, we argue
that neutral theory can be useful to understand the impact of dispersal
on community assembly in landscapes of various complexities. Our main
objective is to use neutral theory to stress the importance of complex landscape
network structure on the distribution of diversity. We refer to the landscape
organization as a "spatial contingency" that could potentially affect the
coexistence mechanisms at play. We briefly review the main approaches to
describe spatial networks and describe three simple toy models of metacommunity
dynamics. We take this opportunity to review their assumptions and main
predictions. We then conduct simple simulations of these models to reveal with
simple examples the impact of spatial network structure on diversity
distribution. The simulation results show that competitive interactions buffer
the potential impact of landscape structure. The strongest centrality-species
richness relationship was observed for the patch dynamics, a model without any
interactions. On the other hand, strong and unequal competitive interactions
minimized the effect of centrality. We conclude that the neutral model is thus a
useful tool to understand the joint effects of dispersal and community
interactions. Our analysis shows that ecologists must now integrate more
realistic landscapes when analyzing community assembly from a metacommunity
perspective.
\newpage

@@

Neutral theory has also been proposed as a useful tool to understand and predict
some aspects of community dynamics. It links to an old philosophical debate
between realism and instrumentalism \parencite{Wennekes2012}. Because every
ecological model is a simplification of reality, scientists have to subjectively
decide the level of details they put in, leaving out some elements they consider
unimportant. The realist perspective requires all assumptions of the theory
to be true, while the utility of the theory is more important to
instrumentalism. The utilitarian value of a theory could either be for
understanding or for prediction (another old philosophical debate, see
\textcite{Shmueli2010}). Obviously neutral theory could only be instrumental.
The question then is if such a 'general, large-scale, but vague' theory
\parencite{Wennekes2012} is a satisfying approximation.

@@

To a vast extent, these relationships are preserved when looking at the
geodesic distances (Fig. 4), i.e. along how many edges should one travel to
connect two patchs. Interestingly enough, the distance-dissimilarity
relationship for the neutral model is markedly hump-shaped, with sites being at
a medium distance having the maximal dissimilarity.

@@

The network approach to the study of spatially explicit landscapes is a major
advancement in metacommunity ecology. It is a first step to make the concept
operational because it accounts for more realistic landscape structures and
dispersal kernels. It makes a significant departure to simple island-mainland or
global dispersal approaches used previously (e.g. \parencite{Tilman1994a,
Hubbell2001, Mouquet2002}). But dispersal is also spatially explicit in a
lattice model and it does not make the landscape more realistic. We believe the
fundamental contribution of this approach is the consideration of spatial
heterogeneity of dispersal. In agreement with previous theoretical
\parencite{Economo2011, Desjardins2012b} and experimemental studies
\parencite{Carrara2012}, the simulations show that the degree centrality has a
significant impact on $\alpha$ diversity. Central nodes might also contribute
more to maintain $\gamma$ diversity, as they are essential for species to spread
throughout the landscape. Such nodes could be potentially quantified as keystone
for the metacommunity \parencite{Mouquet2013}. Interestingly, but not
surprisingly, this effect is weaker with species sorting dynamics. We could even
hypothesize it will vanish in the extreme case of niche differentiation (with
low overlap for instance) and low mass effect. In this particular case, the
neutral versus niche comparison therefore illustrates that very strong unequal
competitive interactions could overwhelm the impact of dispersal.
